### 2014-09-02 writeups/context-dependence.tex

The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
The set of all transition triples, $\calT$, determines the time evolution of the process.
\todo{I'd suggest being more formal and defining a name for the class of processes you describe, and saying what it consists of.}

@@

\paragraph{The generator matrix}
For clarity, we describe how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.

>>>>

The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
The set of all transition triples, $\calT$, determines the time evolution of this process.

@@

\paragraph{The generator matrix}
For clarity, we describe how the set of transition rates on patterns determines the transition rate matrix for complete sequences, that is,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x \in \calS^n$ jumps to state $y \in \calS^n$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h > 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h,\ \text{and} \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$,
and for those positions $x$ matches with $u$ while $y$ matches with $v$.
